Forecasting the 2024 U.S. Presidential Election

### 2024-03-19 paper/uselectionforecast2024.qmd
FINAL EDITS!

The U.S. Presidential election will take place on Tuesday, November 5th, 2024, as America faces unprecedented levels of political polarization and diminished levels of trust in democratic institutions. As of March 11, 2024, former President Trump has 1078 out of the 1215 delegates needed to win the Republican nomination and President Biden has 1872 out of the 1968 delegates needed to win the Democratic nomination, therefore this election will feature the same candidates as the 2020 election. President Joe Biden will seek a second term and former President Donald Trump will try to become the second president to serve two non-consecutive terms. The only President to serve two non-consecutive terms was Grover Cleveland from 1885 to 1889 and 1893 to 1897 [@citeWaxman]. The 2024 election will define the future of women’s rights, LGBTQ+ rights, the separation of church and state, immigration policy, and foreign policy in America. America is at a turning point: Trump and Biden have very different ideologies, policy ideas, and visions for the future of the nation. Former President Trump has celebrated that he was “able to kill Roe v. Wade” [@citeKapur]. With 2022 Supreme Court decision to overturn Roe v. Wade, there will be nothing to stop Trump from signing a bill to restrict access to, or even criminalize, abortion, contraception, or IVF across the United States. In February 2024, Trump stated that he “would encourage” the Putin regime to do “whatever the hell they want” to NATO allies who don’t meet their defense spending targets [@citeSullivan]. On the contrary, President Biden has stood for women’s rights, protecting same-sex marriage, pro-immigration policies, and with Ukraine in their fight for freedom following Russia’s invasion. The January 6th, 2021, insurrection at the U.S. Capitol was a defining moment in the current era of political polarization in America. Trump refused to acknowledge his loss to Joe Biden in the 2020 U.S. presidential election. In August 2020, Trump claimed that the only way he was going to lose the election was “if the election is rigged” [@citeAxelrod]. Since November 2020, he has falsely claimed that the election was fraudulent, and that Biden stole the election from him. On January 6, 2021, an angry pro-Trump mob stormed the U.S. Capitol in an attempt to "stop the steal," or in other words, to stop the joint session of congress from certifying Biden’s electoral college victory [@citeLonsdorf]. It is unclear whether former President Trump will concede if he is defeated again by President Biden in the 2024 election.

Our survey data set was provided by the Polarization Research Lab. The Polarization Research Lab was formed by a group of researchers to increase understanding of where partisan animosity comes from and what can be done to help [@americaspoliticalpulse]. Our post-stratification data set was provided by the Integrated Public Use Microdata Series (IPUMS). IPUMS is an organization which lays out survey and census data with the help of 105 statistical agencies [@citeIPUMS]. We used the data from the 2022 American Community Survey (ACS) for this report [@citeIPUMS]. We use multi-level regression with post-stratification (MRP) to predict the results of the 2024 U.S. Presidential Election. MRP uses a smaller survey data set to fit a model to predict vote preference. We built a logistic regression model, using sex, age, race, hispanic, highest level of education, and whether the respondent lives in an urban or rural area as predictors to estimate support for President Biden. We fit this model using our survey data set, and apply it to our post-stratification data set to predict the popular vote and electoral college results of the 2024 election.

@@

We used R [@citeR], and the `tidyverse` [@citetidyverse], `janitor` [@citejanitor], `ggplot` [@citeggplot], `rstanarm` [@citerstanarm], and `arrow` [@citearrow] packages throughout our analtsis to clean our survey and post-stratification data sets, fit our model, apply our model, and create our data visualizations.

@@

Our surey data set contains zero respondents from Vermont, so it is not shown @fig-electoral-map-survey and the number of electoral colleges in @tbl-surveyresults adds up to 535 instead of 538 because Vermont's 3 electoral votes were not included. The absence of any respondent from Vermont and the fact that there are very few responses from Kansas, Hawaii, Rhode Island, and Wyoming may affect the accuracy of this prediction. This is a significant weakness, and we will discuss it in more depth, as well as how it contributes to the weaknesses and limitations of our model in @sec-weaknesses-limitations.

@@

The sex, race, educ, and metro variables were cleaned and re-factored to mirror our survey data and to allow us to apply our model to our post-stratification data. @fig-demographics1 shows the demographic statistics of our survey and post-stratification data sets. The steps and code for producing @fig-demographics1 were obtained from @citeMatthew. We also created a `race_hispanic`, which is equal to either "hispanic" or "not hispanic" to ensure that our survey and post-stratification data sets would align, because this was a separate variable from race in our post-stratification data. The data cleaning steps used to create the `urban` variable to make our post-stratification data set mirror our survey data set are outlined in @sec-datacleaning.

@@

We can see that there are some slight differences in the proportions of respondents in the survey and post-stratification data from Alabama, California, Florida, Kansas, Maryland, Massachussets, Michigan, New York, Nebraska, and Virginia. None of the voters in our survey data are from Vermont, but 0.198% of the voters in our post-stratification data are. Therefore, the results from applying our model to our post-stratification data will include a prediction for Vermont. @fig-stateprops shows that there is a comparable distribution of voters across U.S. states in the survey and post-stratification data. However, the survey data over represents Florida, Michigan, New York, and Virginia.

@@

@fig-electoralmappoststrat illustrates the same information as @fig-estimatedpropsupport, but on an electoral map. We used the steps and code provided by @citeMatthew and the `statebins` package [@statebins] to produce @fig-electoralmappoststrat. This map shows the mean prediction for the proportion of voters in each state prefer Joe Biden over Donald Trump. We see that California (CA), Maryland (MD), and New Jersey (NJ) are dark blue, indicating strong support for Joe Biden in those states. Wyoming (WY), South Dakota (SD), North Dakota (ND), and Montana (MT) are red, which indicates support for Donald Trump. In addition to this, Texas and Florida are blue in @fig-electoralmappoststrat.The Democrats have not won Florida since former President Barack Obama won when he was re-elected in 2012. Donald Trump carried Florida in both 2016 and 2020, so flipping it in 2024 would be huge for President Biden. The Republican presidential nominee has carried the state of Texas in every presidential election since 1976 [@citeCNN], so flipping it in 2024 would give Biden an excellent chance of winning re-election.

@@

@fig-stanareyouokay-2 shows the trace plots for our predictors. We can see that the lines bounce, but remain horizontal and have a lot of overlap with each other [@tswd]. This suggest that the distribution behaves as expected, and we do not need to re-run our model with fewer predictors or different priors [@tswd].

>>>>

The U.S. Presidential election will take place on Tuesday, November 5th, 2024, as America faces unprecedented levels of political polarization and diminished levels of trust in democratic institutions. As of March 11, 2024, former President Trump has 1078 out of the 1215 delegates needed to win the Republican nomination and President Biden has 1872 out of the 1968 delegates needed to win the Democratic nomination, therefore this election will feature the same candidates as the 2020 election. President Joe Biden will seek a second term and former President Donald Trump will try to become the second president to serve two non-consecutive terms.

Our survey data set was provided by the Polarization Research Lab. The Polarization Research Lab was formed by a group of researchers to increase understanding of where partisan animosity comes from and what can be done to help [@americaspoliticalpulse]. Our post-stratification data set was provided by the Integrated Public Use Microdata Series (IPUMS). IPUMS is an organization which lays out survey and census data with the help of 105 statistical agencies [@citeIPUMS]. We used the data from the 2022 American Community Survey (ACS) for this report [@citeIPUMS]. We use multi-level regression with post-stratification (MRP) to predict the results of the 2024 U.S. Presidential Election. MRP uses a smaller survey data set to fit a model to predict vote preference. We built a logistic regression model, using sex, age, race, Hispanic, highest level of education, and whether the respondent lives in an urban or rural area as predictors to estimate support for President Biden. We fit this model using our survey data set, and apply it to our post-stratification data set to predict the popular vote and electoral college results of the 2024 election.

@@

We used R [@citeR], and the `tidyverse` [@citetidyverse], `janitor` [@citejanitor], `ggplot` [@citeggplot], `rstanarm` [@citerstanarm], and `arrow` [@citearrow] packages throughout our analysis to clean our survey and post-stratification data sets, fit our model, apply our model, and create our data visualizations.

@@

Our survey data set contains zero respondents from Vermont, so it is not shown @fig-electoral-map-survey and the number of electoral colleges in @tbl-surveyresults adds up to 535 instead of 538 because Vermont's 3 electoral votes were not included. The absence of any respondent from Vermont and the fact that there are very few responses from Kansas, Hawaii, Rhode Island, and Wyoming may affect the accuracy of this prediction. This is a significant weakness, and we will discuss it in more depth, as well as how it contributes to the weaknesses and limitations of our model in @sec-weaknesses-limitations.

@@

The sex, race, educ, and metro variables were cleaned and re-factored to mirror our survey data and to allow us to apply our model to our post-stratification data. @fig-demographics1 shows the demographic statistics of our survey and post-stratification data sets. The steps and code for producing @fig-demographics1 were obtained from @citeMatthew. We also created a `race_hispanic`, which is equal to either "Hispanic" or "not Hispanic" to ensure that our survey and post-stratification data sets would align, because this was a separate variable from race in our post-stratification data. The data cleaning steps used to create the `urban` variable to make our post-stratification data set mirror our survey data set are outlined in @sec-datacleaning.

@@

We can see that there are some slight differences in the proportions of respondents in the survey and post-stratification data from Alabama, California, Florida, Kansas, Maryland, Massachussetts, Michigan, New York, Nebraska, and Virginia. None of the voters in our survey data are from Vermont, but 0.198% of the voters in our post-stratification data are. Therefore, the results from applying our model to our post-stratification data will include a prediction for Vermont. @fig-stateprops shows that there is a comparable distribution of voters across U.S. states in the survey and post-stratification data. However, the survey data over represents Florida, Michigan, New York, and Virginia.

@@

@fig-electoralmappoststrat illustrates the same information as @fig-estimatedpropsupport, but on an electoral map. We used the steps and code provided by @citeMatthew and the `statebins` package [@statebins] to produce @fig-electoralmappoststrat. This map shows the mean prediction for the proportion of voters in each state prefer Joe Biden over Donald Trump. We see that California (CA), Maryland (MD), and New Jersey (NJ) are dark blue, indicating strong support for Joe Biden in those states. Wyoming (WY), South Dakota (SD), North Dakota (ND), and Montana (MT) are red, which indicates support for Donald Trump. In addition to this, Texas and Florida are blue in @fig-electoralmappoststrat. The Democrats have not won Florida since former President Barack Obama won when he was re-elected in 2012. Donald Trump carried Florida in both 2016 and 2020, so flipping it in 2024 would be huge for President Biden. The Republican presidential nominee has carried the state of Texas in every presidential election since 1976 [@citeCNN], so flipping it in 2024 would give Biden an excellent chance of winning re-election.

@@

@fig-stanareyouokay-2 shows the trace plots for our predictors. We can see that the lines bounce, but remain horizontal and have a lot of overlap with each other [@tswd]. This suggests that the distribution behaves as expected, and we do not need to re-run our model with fewer predictors or different priors [@tswd].
